Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-07-15 paper/paper.tex
WIP on paper; have a draft model section now

% Note this is the SfN draft abstract, and will need rewriting.
\emph{In the classic Chemoaffinity theory, the retinotectal axon projection is
thought to use pairs of orthogonal signalling gradients in the retina to
specify the eventual location of synapses made on the surface of the
tectum/superior colliculus. Similar orthogonal gradients in the tectum provide
a coordinate system which allows the axons to match their prespecified
destination with the correct location. Although the Ephrins have been shown to
guide axons toward their destination, there has yet to emerge a complete
account of the local interactions which halt the axonal growth cones in the
correct locations to recreate the topography of the retinal cells. The model
of Simpson & Goodhill (2011) provides an account of the basic topographic
arrangement of cells on the tectum, as well as reproducing well known surgical
and genetic manipulation experiments. However, it suffers from the absence of
a local chemotactic guidance mechanism. Instead, each agent in their model is
given instantaneous knowledge of the vector that would move it toward its pre
specified destination. In addition to the globally supervised chemoaffinity
term, Simpson & Goodhill introduced a competitive interaction for space
between growth cone agents and a receptor-ligand axon-axon interaction in
order to account for the full set of experimental manipulations. Here, we
propose the replacement of the chemoaffinity term with a gradient following
model consisting of axonal growth cone agents which carry receptor molecule
expression determined by their soma's location of origin on the retina. Growth
cones move on the simulated tectum guided by two pairs of opposing, orthogonal
signalling molecules representing the Ephrin ligands. We show that with only
the chemoaffinity term and a receptor-ligand based axon-axon interaction term
(meaning that all growth cone interactions are by receptor-ligand signalling),
a full range of experimental manipulations to the retinotectal system can be
reproduced. Furthermore, we show that the observation that competition is not
and essential requirement for axons to find their way (Gosse et al., 2008) is
also accounted for by the model, due to the opposing influences of signalling
gradient pairs. Finally, we demonstrate that, assuming exponentially varying
receptor expression in the retina, ligand expression should either be
exponential if the receptor-ligand signal induces repulsion (i.e. gradient
descent) or logarithmic if the signal induces attraction (gradient
ascent). Thus, we find that a model analogous to the one we presented in James
et al (2020) that accounts for murine barrel patterning is also a candidate
mechanism for the arrangement of the more continuous retinotectal system.}

@@

There have been many approaches to modelling the organisation of the
retino-tectal projection. Many of these models are wholly or partially
phenomenological.

@@

Gradient based models: \citet{nakomoto_topographically_1996}

@@

Our model was inspired by the agent-based model of \cite{simpson_simple_2011}
and, in common with that work, consists of agents representing axonal growth
cones, originating from a square retina and moving upon a square tectum.
%
\citet{simpson_simple_2011} models three contributions to the movement of the growth
cones; chemoaffinity, competition for space and axon-axon interactions. The
first two of these are modelled without reference to a particular mechanism;
each agent is provided with globally-supervised information about its
instantaneous location and its target location, allowing it to move along a
vector towards the target. Competition is implemented as a movement of
constant distance along the vector between two agents if two agents are within
a threshold distance of one another. Interaction is signal-induced repulsion,
which is triggered if two agents are both within the threshold distance of
each other and have compatible EphA receptor expression.

@@

Each retinal ganglion cell projects $n=8$ growth cone agents which carry a set
of 4 receptors, $r$, and 4 ligands, $l$, expressed at levels determined by the
cell's location on the retina. The expression levels of the receptors and
ligands vary with respect to the cell's position on the retinal surface. Each
receptor or ligand varies with respect only to one dimension.  Receptor
expression gradients are arranged in orthogonal pairs with the gradient of
$r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to
$r_0$, is orthogonal to $r_3$. Ligands are also arranged in orthogonal pairs,
with $l_0$ opposing $r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$
orthogonal to $l_0$ and opposing $r_1$.

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
The tectum expresses the same sets of receptors and ligands, also in
orthogonal pairs of gradients. In this work, we did not consider the effect of
reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
X). Correspondingly, we leave open the possibility that ligand expression may
be modelled by exponential, linear or logarithmic functions.

While we do not explicitly name $r_0$, $r_1$,
etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ includes EphA,
EphB, Ryk \citep{schmitt_wntryk_2006} and
Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $l$ includes the
ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and
RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a
role in retino-tectal map formation.

@@

\begin{figure}
\includegraphics[width=\linewidth]{./images/expressions_fig.png}
\caption{Retinal and tectal gene expression.
%
\textbf{A} Retinal receptor expression. Four receptors are expressed and
displayed here in two dual-colour maps. $r_0$ and $r_1$ are red and
green;$r_2$ and $r_3$ are cyan and magenta. $r_0$ increases as an
exponential function of $x$; $r_1$ increases with an exponential function of
$y$. $r_2$ and $r_3$ have the opposite sense. The expression is modelled with
exponential functions because there is ample evidence that shows that retinal
receptor expressions follow this form.
%
\textbf{B} Retinal ligand expression. May not be included.
%
\textbf{C} Tectal receptor expression. May not be included.
%
\textbf{D} Tectal ligand expression. $l_0$ and $l_1$ are red and green; $l_2$
and $l_3$ are cyan and magenta. These expressions are shown in a linear form,
as there is less evidence to determine whether tectal ligand expression is
exponential, linear or logarithmic in form.
}
\label{f:1}
\end{figure}

\subsection{Signalling}

In this model, the signal transmitted when a ligand binds to a receptor on a
growth cone can lead to one of two effects; the signal may induce an
attraction towards the ligand-expression region or a repulsion away from it.
%
For axon-tectum interactions, we implement attraction as a climbing of the
tectal ligand expression gradient and repulsion as gradient descent.
%
To determine the strength of the effect we assume a purely linear receptor
binding model, and set the movement of the growth cone to be $\sum_i \pm
r_i \times \grad l_i$.

For axon-axon interactions, attraction causes a movement along a vector
between two axons causing them to move closer together; repulsion cause the
opposite effect.

%This is where I really want to say ``the only interactions in the model are
%via forward (ligand-activates-receptor) and reverse
%(receptor-activates-ligand) signalling''. Something like:

%Interactions in the model are by receptor-ligand interactions only. Forward
%interactions are those in which the ligand activates the receptor, delivering
%a signal into the receptor-expressing cell. Reverse signals occur when a
%membrane-bound ligand is activated~\citep{wu_role_2019}.

>>>>

% Note this is the SfN draft abstract, and will need rewriting.
\emph{In the classic Chemoaffinity theory, the retinotectal axon projection is
thought to use pairs of orthogonal signalling gradients in the retina to
specify the eventual location of synapses made on the surface of the
tectum/superior colliculus. Similar orthogonal gradients in the tectum provide
a coordinate system which allows the axons to match their prespecified
destination with the correct location. Although the Ephrins have been shown to
guide axons toward their destination, there has yet to emerge a complete
account of the local interactions which halt the axonal growth cones in the
correct locations to recreate the topography of the retinal cells. The model
of \citet{simpson_simple_2011} provides an account of the basic topographic
arrangement of cells on the tectum, as well as reproducing well known surgical
and genetic manipulation experiments. However, it suffers from the absence of
a local chemotactic guidance mechanism. Instead, each agent in their model is
given instantaneous knowledge of the vector that would move it toward its pre
specified destination. In addition to the globally supervised chemoaffinity
term, \citet{simpson_simple_2011} introduced a competitive interaction for
space between growth cone agents and a receptor-ligand axon-axon interaction
in order to account for the full set of experimental manipulations. Here, we
propose the replacement of the chemoaffinity term with a gradient following
model consisting of axonal growth cone agents which carry receptor molecule
expression determined by their soma's location of origin on the retina. Growth
cones move on the simulated tectum guided by two pairs of opposing, orthogonal
signalling molecules representing the Ephrin ligands. We show that with only
the chemoaffinity term and a receptor-ligand based axon-axon interaction term
(meaning that all growth cone interactions are by receptor-ligand signalling),
a full range of experimental manipulations to the retinotectal system can be
reproduced. Furthermore, we show that the observation that competition is not
and essential requirement for axons to find their
way \citep{gosse_retinotopic_2008} is also accounted for by the model, due to
the opposing influences of signalling gradient pairs. Finally, we demonstrate
that, assuming exponentially varying receptor expression in the retina, ligand
expression should either be exponential if the receptor-ligand signal induces
repulsion (i.e. gradient descent) or logarithmic if the signal induces
attraction (gradient ascent). Thus, we find that a model analogous to the one
we presented in \citet{james_modelling_2020} that accounts for murine barrel
patterning is also a candidate mechanism for the arrangement of the more
continuous retinotectal system.}

@@

There have been many approaches to modelling the organisation of the
retinotectal projection. Many of these models are wholly or partially
phenomenological.

@@

Gradient based models: \citet{nakamoto_topographically_1996}

@@

Our model was inspired by the agent-based model of \cite{simpson_simple_2011}
and, in common with that work, consists of agents representing axonal growth
cones, originating from a square retina and moving upon a square tectum.
%
\citet{simpson_simple_2011} model three contributions to the movement of the growth
cones; chemoaffinity, competition for space and axon-axon interactions. The
first two of these are modelled without reference to a particular mechanism;
each agent is provided with globally-supervised information about its
instantaneous location and its target location, allowing it to move along a
vector towards the target. Competition is implemented as a movement of
constant distance along the vector between two agents if those two agents are
within a threshold distance of one another. Interaction is signal-induced
repulsion, which is triggered if two agents are both within the threshold
distance of each other and have compatible EphA receptor expression.

@@

\begin{figure}
\includegraphics[width=\linewidth]{./images/expressions_fig.png}
\caption{Simulated retinal and tectal signalling molecule expression. Four
receptor types are expressed, and four corresponding ligand types. Receptors
of type $i$ are activated by ligands of type $i$.
%
\textbf{A} Retinal receptor expression, $r_i$. Four receptors are expressed and
displayed here in two dual-colour maps. $r_0$ and $r_1$ are red and blue;
$r_2$ and $r_3$ are cyan and magenta. $r_0$ increases as an exponential
function of $-x$; $r_1$ increases with an exponential function of $-y$. $r_2$
and $r_3$ have the opposite sense. Retinal receptor expression is modelled
with exponential functions because there is ample evidence to support this
form. $N$: nasal, $T$: temporal, $D$: dorsal, $V$: ventral.
%
\textbf{B} Retinal ligand expression, $l_i$, is expressed in complimentary
gradients \citep{hornberger_modulation_1999}.
%
\textbf{C} Tectal ligand expression $L_i$. Tectal ligand expression is shown in a linear form,
as there is less evidence to determine whether ligand expression is
exponential, linear or logarithmic on the tectum.
$L$: lateral, $M$: medial, $C$: caudal, $R$: rostral.
}
\label{f:1}
\end{figure}

Each retinal ganglion cell projects $n=8$ growth cone agents (referred to
as \emph{branches}) which carry a set of 4 receptors, $r_i$, and 4 ligands, $l_i$,
expressed at levels determined by the cell's location on the retina. The
expression levels of the receptors and ligands vary with respect to the cell's
position on the retinal surface. Each receptor or ligand varies with respect
only to one dimension.  Receptor expression gradients are arranged in
orthogonal pairs with the gradient of $r_0$ being orthogonal to that of
$r_1$. $r_2$, whose gradient is opposite to $r_0$, is orthogonal to
$r_3$. Ligands are also arranged in orthogonal pairs, with $l_0$ opposing
$r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$ orthogonal to $l_0$ and
opposing $r_1$.

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
The simulated tectum expresses the same types of receptors and ligands, also
in orthogonal pairs of gradients. In this work, we did not consider the effect
of reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression, $L$. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
X). Correspondingly, we leave open the possibility that tectal ligand
expression may be modelled by exponential, linear or logarithmic functions.

While we do not explicitly name $r_0$, $r_1$,
etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ includes EphA,
EphB, Ryk \citep{schmitt_wntryk_2006} and
Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $l$ and $L$
include the ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and
RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a
role in retinotectal map formation.

@@

\subsection{Chemotaxis}

In this model, the signal transmitted when a ligand binds to a receptor on a
growth cone (forward signalling) can lead to one of two effects; the
signal may induce an attraction towards the ligand-expressing region or a
repulsion away from it.
%
For axon-tectum interactions, we implemented attraction as a climbing of the
tectal ligand expression gradient and repulsion as gradient descent.
%
To determine the strength of the effect we assumed a purely linear receptor
binding model, and set the chemotactic movement vector of the growth cone $b$ at
location $\mathbf{x}_b$ on the tectum to be
%
\begin{equation}
\mathbf{G}_b = \sum_i^N F_i\,r_{i,b} \nabla L_i(\mathbf{x}_b)
\end{equation}
%
where $r_{i,b}$ is the receptor expression on axon branch $b$ for
ligand-receptor pair $i$, $\nabla L_i$ is the gradient of expression of ligand
$i$ on the tectum and $F_i$ denotes the direction of the interaction induced
when a molecule of ligand $i$ binds to a receptor $i$ molecule. $F_i$ takes
the value -1 for a repulsive interaction or 1 for an attractive interaction.

\subsection{Axon-axon interactions}

For axon-axon interactions, repulsion causes a movement of branch $b$ along a
unit vector, $\hat{\mathbf{x}}_{kb}$, from branch $k$ to branch $b$ causing
them to move further apart; attraction causes the opposite movement. The
interaction, $\mathbf{I}_b$, acting on axon branch $b$, is given by
%
\begin{equation}
\mathbf{I}_b = \frac{1}{|B_b|} \sum_k \hat{\mathbf{x}}_{kb}\,Q(d_{kb})
\end{equation}
%
where $Q(d_{kb})$ is the signalling strength between two growth cones of
radius $r$ a distance $d_{kb}$ from one another, given by
%
\begin{equation}
Q(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r \\
     0 & d_{kb} > 2r
     \end{cases}
\end{equation}
%
The sign of $Q$ (which is dependent on the values of $F_i$) determines whether
the interaction, $\mathbf{I}_b$, is repulsive or attractive. $l_{i,k}$ is the
expression of ligand $i$ on branch $k$.

\subsection{Movement}

At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, of each
branch, $b$, is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + m_2 \mathbf{I}_b
\end{equation}
%
evaluated at time $t$.
